%!TEX root = ../dissertation.tex
% the acknowledgments section
\singlespacing

First and foremost, I would like to express my deepest gratitude to my supervisor, Johan Hoorn, for his unwavering support and guidance throughout my graduate studies. His expertise, dedication, and encouragement have been invaluable to me.

I would also like to thank the members of my dissertation committee for their time, feedback, and insights. Their constructive comments and suggestions have greatly improved the quality of my work.

Special recognition goes to all the individuals who have contributed directly to this study. I would like to extend my heartfelt thanks to Vicky and Marty, for their collaboration on the studies that are part of this thesis. Their collaborative efforts have not only enriched the quality of this work but also made the research journey more rewarding and enjoyable. Additionally, I am grateful to the participants who generously shared their experiences and perspectives, as well as mentors and colleagues whose invaluable insights and expertise have enriched my research.

I would also like to extend my gratitude to Derek Lomas and Willem van der Maden from TU Delft for their generosity in providing me access to GPT. Their support has enabled me to carry out my research and complete my thesis successfully.

Special thanks go to Mikey and Tesfa for their invaluable help in proofreading my thesis. Their careful attention to detail and suggestions have helped improve the clarity and coherence of my writing.

In addition to my family's love which has sustained me during this journey, I am grateful to my friends who have been my cheerleaders and confidants throughout the process. A special mention goes to Kay and Luci for taking care of my basic necessities such as food, coffee, and for reminding me to sleep when work consumed me.  Their encouragement, thoughtfulness, and camaraderie have made this experience more meaningful and enjoyable.

Lastly, I would like to express my appreciation for everyone else who has played a role in my academic journey—your support, guidance, and belief in me have made a significant impact on my life. Thank you all for your contributions and support.

